maze-dataset

### 2025-03-30 docs/paper/paper.tex
wip, mostly cleaning up

For interpretability and behavioral research, algorithmic tasks offer
benefits by allowing systematic data generation and task decomposition,
as well as simplifying the process of circuit discovery
(\protect\hyperlink{ref-interpretability-survery}{Räuker et al., 2023}).
Although mazes are well suited for these investigations, we have found
that existing maze generation packages
(\protect\hyperlink{ref-cobbe2019procgen}{Cobbe et al., 2019};
\protect\hyperlink{ref-gh_Ehsan_2022}{Ehsan, 2022};
\protect\hyperlink{ref-harriesMazeExplorerCustomisable3D2019}{Harries et
al., n.d.}; \protect\hyperlink{ref-gh_Nemeth_2019}{Németh, 2019};
\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}) do not support flexible maze generation
algorithms that provide fine-grained control of generation parameters
and the ability to easily transform between multiple representations of
the mazes (Images, Textual, Tokenized) for training and testing models.

@@

\begin{itemize}
\item
  Most prior works provide mazes in some kind of image or raster format,
  which is not suitable for training autoregressive text-based
  transformer models -- a key usage case this work seeks to enable.
  However, we still provide a variety of similar output formats:

  \begin{itemize}
  \tightlist
  \item
    we also include the \texttt{RasterizedMazeDataset} class, utilizing
    \texttt{as\_pixels()}, in our codebase, which can exactly mimic the
    outputs provided in
    \texttt{easy-to-hard-data}(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild,
    Borgnia, Gupta, Bansal, et al., 2021}) and can be configured to be
    similar to the outputs of
    (\protect\hyperlink{ref-gh_Nemeth_2019}{Németh, 2019}).
  \item
    Our \texttt{as\_ascii()} method provides a format similar to that
    used in Oppenheim
    (\protect\hyperlink{ref-gh-oppenheimj2018maze}{2018}).
  \item
    Our \texttt{MazePlot} class provides a feature-rich plotting utility
    with support for multiple paths, heatmaps over positions, and more.
    This is similar to the outputs of Ehsan
    (\protect\hyperlink{ref-gh_Ehsan_2022}{2022})
  \end{itemize}
\item
  The text format provided by \texttt{SolvedMaze(...).as\_tokens()} is
  similar to that of (\protect\hyperlink{ref-eval-LLM-graphs}{Liu \& Wu,
  2023}), but provides over 5.8 million unique formats for converting
  mazes to a text stream. We maintain a single underlying format,
  meaning that the same maze can be turned into a variety of text
  streams to assess how the precise format of the text stream affects
  the model.
\item
  For rigorous investigations of the response of a model to various
  distributional shifts, preserving metadata about the generation
  algorithm with the dataset itself is essential. To this end, our
  package efficiently stores the dataset along with its metadata in a
  single human-readable file (\protect\hyperlink{ref-zanj}{M.
  Ivanitskiy, n.d.}). This metadata is loaded when the dataset is
  retrieved from disk and makes it simple to understand how exactly each
  maze was generated. As far as we are aware, no existing packages do
  this reliably.
\item
  Storing mazes as images is not only difficult to work with, but also
  inefficient. Directly storing adjacency matrices is also inefficient
  as subgraphs of the lattice are sparse. Storing adjacency lists can be
  efficient, but comes with a higher lookup cost and possible high
  comparison cost. We use a simple, efficient representation of mazes
  that is optimized for subgraphs of a \(d\)-dimensional finite lattice
  that we do not believe is used in any existing maze generation
  package. This method is described in detail in
  \autoref{implementation}.
\item
  Our package is easily installable with source code freely available.
  It is extensively tested, type hinted, benchmarked, and documented.
  Many other maze generation packages lack this level of rigor and
  scope, and some (\protect\hyperlink{ref-ayaz2008maze}{Ayaz et al.,
  2008}) appear to simply no longer be accessible.
\end{itemize}

@@

When initializing mazes, further configuration options can be specified
through the
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDataset.from_config}{\texttt{from\_config()}}
factory method as necessary. Options allow for saving/loading existing
datasets instead of regenerating, and parallelization options for
generation. Available maze generation algorithms are static methods of
the \texttt{LatticeMazeGenerators} class and include generation
algorithms based on randomized depth-first search, Wilson's algorithm
(\protect\hyperlink{ref-wilson}{Wilson, 1996}), percolation
(\protect\hyperlink{ref-percolation}{Duminil-Copin, 2017};
\protect\hyperlink{ref-percolation-clustersize}{Fisher \& Essam, 2004}),
Kruskal's algorithm
(\protect\hyperlink{ref-kruskal1956shortest}{Kruskal, 1956a}), and
others.

@@

\begin{figure}[H]
  \centering
  \begin{tabular}{p{2in} p{2in} p{2in}} 
    \hline \\[.5em]
    % algorithms
    \texttt{as\_ascii()} & \texttt{as\_pixels()} & \texttt{MazePlot()} \\[.5em]
    % descriptions
      Simple text format for displaying mazes, useful for debugging in a terminal environment.
      & \texttt{numpy} array of \texttt{dtype=uint8} and shape \texttt{(height, width, 3)}. The last dimension is RGB color.
      & feature-rich plotting utility with support for multiple paths, heatmaps over positions, and more. \\[1em]
    \hline \\
    % examples
      \multicolumn{1}{c}{\begin{minipage}[b]{1.6in}
        \input{figures/outputs-ascii-colored.tex} 
      \end{minipage}}
      & \multicolumn{1}{c}{
        \includegraphics[width=0.25\textwidth]{figures/outputs-pixels.pdf}
      }
      & \multicolumn{1}{c}{
        \includegraphics[width=0.25\textwidth, trim={0 0 -.3cm, -.5cm}, clip]{figures/outputs-mazeplot.pdf}
      } \\[1em]

@@

All output sequences consist of four token regions representing
different features of the maze. These regions are distinguished by color
in \autoref{fig:token-regions}.

\begin{itemize}
\tightlist
\item
  {Adjacency list}: A text representation of the lattice graph
\item
  {Origin}: Starting coordinate
\item
  {Target}: Ending coordinate
\item
  {Path}: Maze solution sequence from the start to the end
\end{itemize}

\begin{figure}
\hypertarget{fig:token-regions}{%
\centering
\includegraphics{figures/outputs-tokens-colored.tex}
\caption{Example text output format with token regions
highlighted.}\label{fig:token-regions}
}
\end{figure}

@@

\hypertarget{success-rate-estimation}{%
\subsubsection{Success Rate Estimation
Algorithm}\label{success-rate-estimation}}

@@

\begin{itemize}
\item
  \textbf{Shifted Sigmoid}: Creates sharp transition at \(x=0.5\)
  \[\sigma_s(x) = (1 + e^{-10^3 \cdot (x-0.5)})^{-1}\]
\item
  \textbf{h-function}: Transition function combining polynomial and
  sigmoid components
  \[h(q,a) = q \cdot (1 - |2q-1|^a) \cdot (1-\sigma_s(q)) + (1-(1-q) \cdot (1 - |2(1-q)-1|^a)) \cdot \sigma_s(q)\]
\item
  \textbf{Amplitude Scaling}: Weight-modulated polynomial
  \[A(q,a,w) = w \cdot (1 - |2q-1|^a)\]
\item
  \textbf{Soft Step}: Identity-like for \(p \approx 0.5\), pushes \(x\)
  to extremes otherwise
  \[\text{soft\_step}(x, p, \alpha, w) = h(x, A(p, \alpha, w))\]
\end{itemize}

Finally, we define

\[
  \text{cfg\_success\_predict\_fn}(\mathbf{x}) = \text{soft\_step}(\text{raw\_val}, x_0, 5, 10)
\]

@@

\begin{figure}
\centering
\includegraphics{figures/ep/ep_deadends_unique.pdf}
\caption{An example of both empirical and predicted success rates as a
function of the percolation probability \(p\) for various maze sizes,
percolation with and without depth first search, and
\texttt{endpoint\_kwargs} requiring that both the start and end be in
unique dead ends.}
\end{figure}

@@

This package was originally built for the needs of the
(\protect\hyperlink{ref-maze-transformer-github}{Michael I. Ivanitskiy
et al., 2023b}) project, which aims to investigate spatial planning and
world models in autoregressive transformer models trained on mazes Spies
et al. (\protect\hyperlink{ref-spies2024causalworldmodels}{2024}). This
project has also adapted itself to be useful for work on understanding
the mechanisms by which recurrent convolutional and implicit networks
(\protect\hyperlink{ref-fung2022jfb}{Fung et al., 2022}) solve mazes
given a rasterized view
(\protect\hyperlink{ref-knutson2024logicalextrapolation}{Knutson et al.,
2024}), and for this we match the output format of
(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}).

@@

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

The
\href{https://github.com/understanding-search/maze-dataset}{\texttt{maze-dataset}}
library (\protect\hyperlink{ref-maze-dataset-github}{Michael I.
Ivanitskiy et al., 2023a}) introduced in this paper provides a flexible
and extensible toolkit for generating, processing, and analyzing maze
datasets. By supporting various procedural generation algorithms and
conversion utilities, it enables the creation of mazes with customizable
properties to suit diverse research needs. Planned improvements to the
\texttt{maze-dataset} include adding more generation algorithms (such as
Prim's algorithm (\protect\hyperlink{ref-dijkstra-prim}{Dijkstra, 1959};
\protect\hyperlink{ref-jarnik-prim}{Jarnık, 1930};
\protect\hyperlink{ref-prim}{Prim, 1957}) and Kruskal's algorithm
(\protect\hyperlink{ref-kruskal}{Kruskal, 1956b}), among others
(\protect\hyperlink{ref-mazegen_analysis}{Gabrovšek, 2019})), adding the
ability to augment a maze with an adjacency list to add ``shortcuts'' to
the maze, and resolving certain limitations detailed in Section
\protect\hyperlink{limitations}{Limitations}.

@@

This work was partially supported by and many of the authors were
brought together by AI Safety Camp and AI Safety Support. This work was
partially funded by National Science Foundation awards DMS-2110745 and
DMS-2309810. We are also grateful to LTFF and FAR Labs for hosting three
of the authors for a Residency Visit, and to various members of FAR's
technical staff for their advice. We thank the Mines Optimization and
Deep Learning group (MODL) for fruitful discussions. We also thank
Michael Rosenberg for recommending the usage of Finite State Transducers
for storing tokenizer validation information.

@@

\leavevmode\vadjust pre{\hypertarget{ref-dijkstra-prim}{}}%
Dijkstra, E. W. (1959). \emph{A note on two problems in connexion with
graphs:({Numerische Mathematik}, 1 (1959), p 269-271)}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-mazegen_analysis}{}}%
Gabrovšek. (2019). Analysis of maze generating algorithms. \emph{{IPSI
Transactions} on {Internet Research}}, \emph{15.1}, 23--30.
\url{http://www.ipsitransactions.org/journals/papers/tir/2019jan/p5.pdf}

@@

\leavevmode\vadjust pre{\hypertarget{ref-jarnik-prim}{}}%
Jarnık, V. (1930). About a certain minimal problem. \emph{Práce Moravské
Prırodovedecké Spolecnosti}, \emph{6}, 57--63.

@@

\leavevmode\vadjust pre{\hypertarget{ref-kruskal1956shortest}{}}%
Kruskal, J. B. (1956a). On the shortest spanning subtree of a graph and
the traveling salesman problem. \emph{Proceedings of the American
Mathematical Society}, \emph{7}(1), 48--50.

\leavevmode\vadjust pre{\hypertarget{ref-kruskal}{}}%
Kruskal, J. B. (1956b). On the shortest spanning subtree of a graph and
the traveling salesman problem. \emph{Proceedings of the American
Mathematical Society}, \emph{7}(1), 48--50.
\url{https://doi.org/10.1090/S0002-9939-1956-0078686-7}

@@

\leavevmode\vadjust pre{\hypertarget{ref-prim}{}}%
Prim, R. C. (1957). Shortest connection networks and some
generalizations. \emph{The Bell System Technical Journal}, \emph{36}(6),
1389--1401.

>>>>

For interpretability and behavioral research, algorithmic tasks offer
benefits by allowing systematic data generation and task decomposition,
as well as simplifying the process of circuit discovery
(\protect\hyperlink{ref-interpretability-survery}{Räuker et al., 2023}).
Although mazes are well suited for these investigations, we have found
that existing maze generation packages
(\protect\hyperlink{ref-cobbe2019procgen}{Cobbe et al., 2019};
\protect\hyperlink{ref-gh_Ehsan_2022}{Ehsan, 2022};
\protect\hyperlink{ref-harriesMazeExplorerCustomisable3D2019}{Harries et
al., n.d.}; \protect\hyperlink{ref-gh_Nemeth_2019}{Németh, 2019};
\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}) do not support flexible maze generation
algorithms that provide fine-grained control of generation parameters
and the ability to easily transform between multiple representations of
the mazes (Raster, Textual, Tokenized) for training and testing models.

@@

\begin{itemize}
\item
  Most prior works provide mazes in some kind of image or raster format,
  and we provide a variety of similar output formats:

  \begin{itemize}
  \tightlist
  \item
    \texttt{RasterizedMazeDataset}, utilizing \texttt{as\_pixels()},
    which can exactly mimic the outputs provided in
    \texttt{easy-to-hard-data}(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild,
    Borgnia, Gupta, Bansal, et al., 2021}) and can be configured to be
    similar to the outputs of Németh
    (\protect\hyperlink{ref-gh_Nemeth_2019}{2019})
  \item
    \texttt{as\_ascii()} provides a format similar to that used in
    (\protect\hyperlink{ref-gh-oppenheimj2018maze}{Oppenheim, 2018};
    \protect\hyperlink{ref-eval-gpt-visual}{Singla, 2023})
  \item
    \texttt{MazePlot} provides a feature-rich plotting utility with
    support for multiple paths, heatmaps over positions, and more. This
    is similar to the outputs of
    (\protect\hyperlink{ref-mazegenerator-net}{Alance AB, 2019};
    \protect\hyperlink{ref-gh_Ehsan_2022}{Ehsan, 2022};
    \protect\hyperlink{ref-mathematica-maze}{Guo et al., 2011};
    \protect\hyperlink{ref-mdl-suite}{Nag, 2020})
  \end{itemize}
\item
  The text format provided by \texttt{SolvedMaze(...).as\_tokens()} is
  similar to that of (\protect\hyperlink{ref-eval-LLM-graphs}{Liu \& Wu,
  2023}), but provides over 5.8 million unique formats for converting
  mazes to a text stream, detailed in
  \autoref{tokenized-output-formats}.
\item
  For rigorous investigations of the response of a model to various
  distributional shifts, preserving metadata about the generation
  algorithm with the dataset itself is essential. To this end, our
  package efficiently stores the dataset along with its metadata in a
  single human-readable file (\protect\hyperlink{ref-zanj}{M.
  Ivanitskiy, n.d.}). As far as we are aware, no existing packages do
  this reliably.
\item
  Storing mazes as images is not only difficult to work with, but also
  inefficient. We use an extremely efficient method detailed in
  \autoref{implementation}.
\item
  Our package is easily installable with source code freely available.
  It is extensively tested, type hinted, benchmarked, and documented.
  Many other maze generation packages lack this level of rigor and
  scope, and some (\protect\hyperlink{ref-ayaz2008maze}{Ayaz et al.,
  2008}) appear to simply no longer be accessible.
\end{itemize}

@@

When initializing mazes, further configuration options can be specified
through the
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDataset.from_config}{\texttt{from\_config()}}
factory method as necessary. Options allow for saving/loading existing
datasets instead of regenerating, and parallelization options for
generation. Available maze generation algorithms are static methods of
the \texttt{LatticeMazeGenerators} class and include generation
algorithms based on randomized depth-first search, Wilson's algorithm
(\protect\hyperlink{ref-wilson}{Wilson, 1996}), percolation
(\protect\hyperlink{ref-percolation}{Duminil-Copin, 2017};
\protect\hyperlink{ref-percolation-clustersize}{Fisher \& Essam, 2004}),
Kruskal's algorithm
(\protect\hyperlink{ref-kruskal1956shortest}{Kruskal, 1956}), and
others.

@@

\begin{figure}[H]
  \centering
  \begin{tabular}{p{1.5in} p{1.5in} p{1.5in}} 
    \hline \\[.5em]
    % algorithms
    \texttt{as\_ascii()} & \texttt{as\_pixels()} & \texttt{MazePlot()} \\[.5em]
    % descriptions
      Simple text format for displaying mazes, useful for debugging in a terminal environment.
      & \texttt{numpy} array of \texttt{dtype=uint8} and shape \texttt{(height, width, 3)}. The last dimension is RGB color.
      & feature-rich plotting utility with support for multiple paths, heatmaps over positions, and more. \\[1em]
    \hline \\
    % examples
      \multicolumn{1}{c}{\begin{minipage}[b]{1.6in}
        \setlength{\baselineskip}{0.9em}
        \input{figures/outputs-ascii-colored.tex} 
      \end{minipage}}
      & \multicolumn{1}{c}{
        \includegraphics[width=0.25\textwidth]{figures/outputs-pixels.pdf}
      }
      & \multicolumn{1}{c}{
        \includegraphics[width=0.25\textwidth, trim={0 0 -.3cm, -.5cm}, clip]{figures/outputs-mazeplot.pdf}
      } \\[1em]

@@

All output sequences consist of four token regions representing
different features of the maze of which we see an example in
\autoref{fig:token-regions}.

\begin{figure} 
  \centering
  \begin{minipage}{5in}
    \footnotesize
    \input{figures/outputs-tokens-colored.tex}
  \end{minipage}
  \caption{
    Example text output format with token regions highlighted.
    \colorbox[RGB]{ 217,210,233 }{Adjacency list}: text representation of the lattice graph,
    \colorbox[RGB]{ 217,234,211 }{Origin}: starting coordinate,
    \colorbox[RGB]{ 234,209,220 }{Target}: ending coordinate,
    \colorbox[RGB]{ 207,226,243 }{Path}: maze solution sequence
  }
  \label{fig:token-regions}
\end{figure}

@@

\hypertarget{success-rate-estimation-algorithm}{%
\subsubsection{Success Rate Estimation
Algorithm}\label{success-rate-estimation-algorithm}}

@@

A shifted sigmoid \(\sigma_s\), amplitude scaling \(A\), and \(h\)
function: \[
  \sigma_s(x) = (1 + e^{-10^3 \cdot (x-0.5)})^{-1}
  \qquad A(q,a,w) = w \cdot (1 - |2q-1|^a)
\] \[
  h(q,a) = q \cdot (1 - |2q-1|^a) \cdot (1-\sigma_s(q)) + (1-(1-q) \cdot (1 - |2(1-q)-1|^a)) \cdot \sigma_s(q)
\]

We combine these to get the \texttt{soft\_step} function, which is
identity-like for \(p \approx 0.5\), and pushes pushes \(x\) to extremes
otherwise. \[
  \text{soft\_step}(x, p, \alpha, w) = h(x, A(p, \alpha, w))
\]

Finally, we define \[
  \text{cfg\_success\_predict\_fn}(\mathbf{x}) = \text{soft\_step}(\text{raw\_val}, x_0, 5, 10)
\]

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth,height=\textheight]{figures/ep/ep_deadends_unique.pdf}
\caption{An example of both empirical and predicted success rates as a
function of the percolation probability \(p\) for various maze sizes,
percolation with and without depth first search, and
\texttt{endpoint\_kwargs} requiring that both the start and end be in
unique dead ends.}
\end{figure}

@@

This package was originally built for the needs of the
(\protect\hyperlink{ref-maze-transformer-github}{Michael I. Ivanitskiy
et al., 2023b}) project, which aims to investigate spatial planning and
world models in autoregressive transformer models trained on mazes
(\protect\hyperlink{ref-ivanitskiy2023structuredworldreps}{Michael
Igorevich Ivanitskiy et al., 2023};
\protect\hyperlink{ref-spies2024causalworldmodels}{Spies et al., 2024}).
This project has also adapted itself to be useful for work on
understanding the mechanisms by which recurrent convolutional and
implicit networks (\protect\hyperlink{ref-fung2022jfb}{Fung et al.,
2022}) solve mazes given a rasterized view
(\protect\hyperlink{ref-knutson2024logicalextrapolation}{Knutson et al.,
2024}), and for this we match the output format of
(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}).

@@

@@

This work was partially supported by and many of the authors were
brought together by AI Safety Camp and AI Safety Support. We would like
to thank our former collaborators at AI Safety Camp and other users and
contributors to the \texttt{maze-dataset} package: Naveen Arunachalam,
Benji Berczi, Guillaume Corlouer, William Edwards, Leon Eshuijs, Chris
Mathwin, Lucia Quirke, Can Rager, Adrians Skapars, Johannes Treutlein,
and Dan Valentine.

This work was partially funded by National Science Foundation awards
DMS-2110745 and DMS-2309810. We are also grateful to LTFF and FAR Labs
for hosting three of the authors for a Residency Visit, and to various
members of FAR's technical staff for their advice.

We thank the Mines Optimization and Deep Learning group (MODL) for
fruitful discussions. We also thank Michael Rosenberg for recommending
the usage of Finite State Transducers for storing tokenizer validation
information.

@@

@@

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-kruskal1956shortest}{}}%
Kruskal, J. B. (1956). On the shortest spanning subtree of a graph and
the traveling salesman problem. \emph{Proceedings of the American
Mathematical Society}, \emph{7}(1), 48--50.

@@
